### 2020-10-16 doc/manuscript/Dow_manuscript.Rmd
intro work

---
title: ''
output: 
  word_document: 
    fig_caption: yes
bibliography: references.bib
csl: global-change-biology.csl #(this will change depending on journal we choose. GCB would be a great choice)
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

In recent decades, an increase in average Spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). The effects of this Spring warming are most apparent when examining leaf phenology shifts, which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many interesting observations. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of Spring warming also has a significant effect on leaf phenology response. Warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016). Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).

Shifts in leaf phenology timing could be accompanied by a change in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). However, the fine-scale responses of temperate deciduous species to a warming Spring are not commonly researched due to the increased effort required. Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review). To date, very little research, has been done on the phenology of stem-growth.

In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in Spring temperature patterns. Timing of intraannual stem-growth is extremely important as it could have profound impacts on ecosystem interactions, as well as dictate periods of highest drought sensitivity of NPP in temperate forests (Foster, 2014; ). For example, if a drought were to occur when trees were at their peak growth rates, it could decrease that year's NPP substantially, and subsequently, reduce the total amount of carbon sequestered, contributing to a negative climate feedback loop (source?). Recent results from a paper by D’Orangeville et al. (in review) have shown that wood-type plays a significant role in the timing of leaf and wood phenology. They observed that ring-porous trees begin growth earlier in the season compared to diffuse-porous trees, but both reached 75% total annual growth around the same time. This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stresses.

^More info on wood-type? move some stuff up from discussion?

While not entirely impossible, it is difficult to record intraannual growth or compare interspecific responses to Spring warming using leaf-phenology or other indirect observations. However, traditional methods of direct observation are labor intensive, nad very time consuming (Source about cell sampling). A solution to this is the use of dendrometer bands. Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale (source?). Using many measurements taken within a year, we can model the growth of individual trees and extract growth milestones and variables for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014). This allows the determination of variable responses of all tree species within a single forest, meaning a better understanding of future growth under projected climate conditions. Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual stem-growth phenology within the ForestGEO plot at SCBI, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth. Correlation between these variables and Spring temperatures over the 9 year study period (2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.

We hypothesized that (1) stem-growth phenology variables would occur earlier in warmer years as compared to cooler years and (2) both wood types would increase total growth and max growth rate in response to warmer Spring temperatures.

@@

The mechanisms controlling activation of cambium reactivation is still debated, as some speculate Auxin, the chemical believed to kick-start the reactivation process, originates in the leaves (Frankenstein et al, 2005). However, the fact that ring-porous species complete 25% growth before the presence of leaf buds indicates there may be another source of Auxin production.

Under the warming regime analyzed here, ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while timing of 75% growth remains statistically unchanged. This implies that the growing season of ring-porous trees expanded due to Spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies. In fact, we found a positive correlation between growing season length in ring-porous trees and Spring temperature, while no such relationship was observed in diffuse-porous trees. The earlier reactivation of cambial cells is potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990). Damage of this nature would then need to be repaired using carbon reserves, which may be in short demand as they are already being used for initial growth.This would negatively impact the plant's ability to maintain cell health, or reduce the ability to respond to invasion of pests or pathogens. Alternatively, the extension of growth season and higher yearly growth totals like we see here, could give ring-porous trees a competitive advantage over their non-responsive diffuse-porous counterparts.

@@

Content to incorporate: N. American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts. Thus, climate change is having less impact (Zohner et al., 2020); Leaf velocity discussion?

>>>>

---
title: ''
output:
  word_document:
    fig_caption: yes
  pdf_document: default
bibliography: references.bib
csl: global-change-biology.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

In recent decades, an increase in average Spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). The effects of this Spring warming are most apparent when examining leaf phenology shifts, which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of this Spring warming also has a significant effect on leaf phenology response. Warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016). Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018). Therefore, timing of seasonal warming could influence when bud burst occurs, and consequently, lead to variations in total CO2 sequestered by deciduous forests.

Shifts in leaf phenology timing could be accompanied by a change in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). However, the fine-scale responses of temperate deciduous tree growth to a warming Spring are not commonly recorded due to the increased effort required (). Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review). Further to this point, a recent publication by D'Orangeville et al () found that leaf and wood phenology traits are dissimilar among trees of different wood types (ring-porous and diffuse-porous). For example, they observed that bud burst was occuring around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.

The significance of wood type in regards to wood and leaf phenology likely implies evolutionary differences and adaptations to local climates where these trees can be found. Ring-porous trees have large xylem cells which are prone to embolism during the winter. These cells are repaired before leaf development, explaining why the 25% yearly growth milestone is reached around the same time as bud burst. Diffuse-porous xylem cells are smaller, and less prone to embolism during the Winter, meaning they do not need to be repaired in order to restore hydraulic functionality before beginning the leaf development process. Results from the paper by D’Orangeville et al. (in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during Spring, while diffuse-porous trees reach these milestones in the Summer, but both reached 75% total annual growth around the same time. --Consequently, the Summer focused diffuse-porous trees are exposed to higher temperatures and drought risk than ring-porous trees.-- If leaf development is sped up when Spring temperatures are warmer, we should seek to understand whether this relationship holds, and how it potentially varies among trees of different wood-types.

Earlier leaf up timings could signal that these trees are shifting their peak growth windows back in response to increasing Spring temperatures, but to date, very little research, has been done on the phenology of stem-growth. In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in Spring temperature patterns. Timing of intraannual stem-growth is extremely important as it could have profound impacts on ecosystem interactions, as well as dictate periods of highest drought sensitivity in temperate forests (Foster, 2014; ). If a drought were to occur when trees were at their peak growth rates, the resulting decrease in that year's NPP would reduce the total amount of carbon sequestered, contributing to a climate feedback loop (source?).

While not entirely impossible, it is difficult to record intraannual growth or compare interspecific responses to Spring warming using leaf-phenology or other indirect observations. However, traditional methods of direct observation are labor intensive, and very time consuming (Source about cell sampling). A solution to this is the use of dendrometer bands. Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale (source?). Using many measurements taken within a year, we can model the growth of individual trees and extract growth milestones and variables for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014). This allows the determination of variable responses of all tree species within a single forest, meaning a better understanding of future growth under projected climate conditions. Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual stem-growth phenology within a forest dynamics plot in the Eastern USA, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth. Correlation between these variables and Spring temperatures over the 9 year study period (2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.

We hypothesized that (1) stem-growth milestones would be reached earlier in warmer years as compared to cooler years and (2) both wood types would increase total growth and max growth rate in response to warmer Spring temperatures.

@@

The mechanisms controlling activation of cambium reactivation is still debated, as some speculate Auxin, the chemical believed to kick-start the reactivation process, originates in the leaves (Frankenstein et al, 2005). However, the fact that ring-porous species complete 25% growth before the presence of leaf buds indicates there may be another source of Auxin production. *READ MORE ABOUT THIS!* The relationship between earlier cambial reactivation and early Spring temperature signals that the mechanism is influenced by temperature?

Under the warming regime analyzed here, ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while timing of 75% growth !!! Changed by constant amount !!! . This implies that the growing season of ring-porous trees expanded due to Spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies. In fact, we found a positive correlation between ring-porous growing season length and Spring temperature, while no such relationship was observed in diffuse-porous trees. The earlier reactivation of cambial cells is potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990). Damage of this nature would then need to be repaired using carbon reserves, which may be in short demand as they are already being used for initial growth.This would negatively impact the plant's ability to maintain cell health, or reduce the ability to respond to invasion of pests or pathogens. Alternatively, the extension of growth season and higher yearly growth totals like we see here, could give ring-porous trees a competitive advantage over their non-responsive diffuse-porous counterparts.

@@

Content to incorporate: N. American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts. Thus, climate change is having less impact (Zohner et al., 2020); Leaf velocity discussion? - - RP trees BB = 25% and 75% leaf development = 50% growth (D'Orangeville). Are DP trees leafing up earlier? If yes, how much earlier? Similar to 25,50,75% DOY retreat?
